---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mthapower

Calculate sample size and post-hoc power of association studies involving mitochondrial DNA haplogroups - Based on Samuels et al. AJHG, 2006. 78(4):713-720. [DOI:10.1086/502682](https://www.ncbi.nlm.nih.gov/pmc/PMC1424681)

## Installation

- From CRAN: 

```{r cran-installation, eval = FALSE}
install.packages("mthapower")
```

- From GitHub:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("aurora-mareviv/mthapower")
```

## Shiny app

- Run in Shinyapps.io: [mtDNA_power_calc](https://aurora.shinyapps.io/mtDNA_power_calc/)
- Run locally from Gist: 

```{r gh-rungist, eval = FALSE}
# install.packages("shiny")
shiny::runGist('5895082')
```

## Examples

### Sample size estimation

- Determine the minimum number of cases (`Ncmin`), required to detect: either a change from `p0` (haplogroup frequency in controls) to `p1` (haplogroup frequency in cases), or a given OR, with a predefined confidence interval, in a study with `Nh` haplogroups.

```{r example, message=FALSE, warning=FALSE}
library(mthapower)
library(dplyr)
mydata <- mthacases(p0=0.445, Nh=11,
                    OR.cas.ctrl=c(2), power=80,
                    sig.level=0.05) # Baudouin study
mydata <- mthacases(p0=0.445, Nh=11,
                    OR.cas.ctrl=c(1.25,1.5,1.75,2,2.25,2.5,2.75,3),
                    power=80, sig.level=0.05)
mydata <- mydata[c(2,6)]
mydata %>%
  knitr::kable()
plot(mydata)
```

### Power estimation

- For a given study size, determine the minimum effect size that can be detected with the desired power and significance level, in a study with `Nh` haplogroups. 

```{r example2a, message=FALSE, warning=FALSE}
# Example 2a:
# library(mthapower)
pow <- mthapower(n.cases=203, p0=0.443, Nh=13, OR.cas.ctrl=2.33, sig.level=0.05)
pow %>%
  knitr::kable()
```

```{r example2b, message=FALSE, warning=FALSE}
# Example 2b:
# Create data frames
pow.H150 <- mthapower(n.cases=seq(50,1000,by=50), p0=0.433, Nh=11,
                      OR.cas.ctrl=1.5, sig.level=0.05)
pow.H175 <- mthapower(n.cases=seq(50,1000,by=50), p0=0.433, Nh=11,
                      OR.cas.ctrl=1.75, sig.level=0.05)
pow.H200 <- mthapower(n.cases=seq(50,1000,by=50), p0=0.433, Nh=11,
                      OR.cas.ctrl=2, sig.level=0.05)
pow.H250 <- mthapower(n.cases=seq(50,1000,by=50), p0=0.433, Nh=11,
                      OR.cas.ctrl=2.5, sig.level=0.05)

# Bind the three data frames:
bindata <- rbind(pow.H150,pow.H175,pow.H200,pow.H250)
# Adds column OR to binded data frame:
bindata$OR <- rep(factor(c(1.50,1.75,2,2.5)),
              times = c(nrow(pow.H150),
                        nrow(pow.H175),
                        nrow(pow.H200),
                        nrow(pow.H250)))
# Create plot:
# install.packages("car")
library(car)
scatterplot(power~ncases | OR, reg.line=FALSE,
            smooth=FALSE, spread=FALSE,
            boxplots=FALSE, span=0.25, by.groups=FALSE,
            data=bindata)
```
